---
title: "Notes on Ch4 - The Ames Housing Data"
author: "N_Lim"
date: "2025-06-25"
output:
  pdf_document: default
  html_document: default
---

Loading the data:
```{r}
library(modeldata)
data(ames)
```

checking the data
```{r}
dim(ames)
```

basic EDA on ames data
```{r}
# Load packages
library(tidymodels)
tidymodels_prefer()
```

Actual start of EDA
```{r}
ggplot(ames, aes(x = Sale_Price)) +
  geom_histogram(bins = 50, col = "white")
```

Observations on the plot:
- data are right-skewed
- there are more expensive houses in the dataset than inexpensive ones
- log-transforming the data first before modeling is a good decision
- log-transformation will also stabilize the variance in the data, making our inference more 'legit'

Plot of log-transformed data:
```{r}
ggplot(ames, aes(x = Sale_Price)) +
  geom_histogram(bins = 50, col = "white") +
  scale_x_log10()
```

While not perfect, the data looks closer to bell curve now...

Proceeding to log-transform the data:
```{r}
ames <- ames |> 
  mutate(Sale_Price = log10(Sale_Price))
```






